```{r include=FALSE, cache=FALSE}
library(R.ComDim)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 5,
  fig.height = 5,
  message = FALSE
)
```

```{r, echo = FALSE, eval = TRUE}
library(R.ComDim)
```

# Create a MultiBlock {#create}

## Option 1: for small MultiBlocks {#option1}
In order to use the ComDim algorithm, the data blocks need to be combined
into a `MultiBlock` object.

`MultiBlock` has 5 fields: `Samples`, `Data`, `Variables`,
`Metadata` and `Batch`. The easiest way to create a `MultiBlock` is by using the
function `MultiBlock()` as below:

```{r create, echo = TRUE, eval = FALSE}
b1 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
b2 = as.data.frame(matrix(rnorm(800),10,80)) # 10 rows and 80 columns
b2[c(2,3,4),c(5,7,8)] <- NA # Making some data missing, just for fun.
rownames(b2) <- LETTERS[5:14] # Adding some sample names
b3 = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                s2 = 1:ncol(b2)))
```

With the code above, a `MultiBlock` containing 2 blocks was built. As shown, the
provided data blocks support the format `matrix` and `data.frame`.

When building a `MultiBlock`, only the fields `Samples`, `Data`, and `Variables`
are mandatory.

```{r create2, echo = TRUE, eval = FALSE}
b4 = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                 s2 = 1:ncol(b2)),
                Batch = list(s1 = rep('Batch1',10),
                             s2 = c(rep('Batch2',5),rep('Batch3',5))))
```

And to create a `MultiBlock` with sample metadata, the easiest way is with the
function `AddMetadata`:

```{r AddMetadata, echo = TRUE, eval = FALSE}
b4 <- AddMetadata(newBlock = matrix(rnorm(100),10,10),
                  metadata = data.frame(x1 = c(rep(0,5),rep(1,5)),
                                        x2 = c(rep(1,3),rep(2,4),rep(3,3))))
```

In `AddMetadata`, newBlock can be a matrix, a data.frame, or a MultiBlock 
object. In the giving example, the metadata contains two variables (`x1` and 
`x2`).


## Option 2: for large MultiBlocks {#option2}

Then, it is possible to create a `MultiBlock` from another preexisting
`MultiBlock`. In this case, the function `BuildMultiBlock()` comes in handy.

```{r create3, echo = TRUE, eval = FALSE}
MB <- BuildMultiBlock(b1,b2,b3,ignore.names = TRUE)
getBlockNames(MB)
```


## Option 3: from SummarizedExperiment or MultiAssayExperiment objects. {#option3}

`MultiBlock`s can also be created from `SummarizedExperiment` or 
`MultiAssayExperiment` objects, which are pretty common format nowadays in
 multi-omics studies. In the same way, `MultiBlocks` can be converted back to 
 `MultiAssayExperiments`.

```{r SummarizedExperiment, echo = TRUE, eval = FALSE}
library(MultiAssayExperiment)
data(miniACC)
library(SummarizedExperiment)
data(airway, package="airway")

MB1 <- SummarizedExperiment2MultiBlock(airway,
                                       colData_samplenames = 'Run',
                                       Batch = NULL)
MB2 <- MultiAssayExperiment2MultiBlock(miniACC,
                                       colData_samplenames = 'patientID',
                                       Batch = NULL)
se2 <-  MultiBlock2MultiAssayExperiment(MB2, MSEmetadata = NULL)
```



## Sample correspondence across blocks {#correspondence}

We define as __sample correspondence across blocks__ to the fact that the sample
 order is maintained across blocks. That is, the first sample from the first 
 block has a correspondence sample in the second block at the same position, and
 so on.
 
To run ComDim **and most multi-omics analysis** in general, sample 
correspondence across blocks is needed. This is normally verified by the user.
For `R.ComDim`, we can use `BuildMultiBlock()` to check for the same 
correspondence and resort samples (and discard) if needed. This process is 
achieved with the option `ignore.names = FALSE` (the default).

```{r correspondence, echo = TRUE, eval = FALSE}
c1 = matrix(1:500,10,50) # 10 rows and 50 columns
c2 = matrix(500:1,10,50) # 10 rows and 50 columns
c3 = matrix(501:1000,10,50) # 10 rows and 50 columns
c4 = matrix(1:1000,20,50) # 20 rows and 50 columns
rownames(c1) <- paste0('c',6:15)
rownames(c2) <- paste0('c',1:10)
rownames(c3) <- paste0('c',10:1)
rownames(c4) <- paste0('c',1:20)
# With ignore.names = FALSE, only common samples across blocks are kept.
# Samples will be resorted if needed.
MB12 <- BuildMultiBlock(c1,c2,ignore.names = FALSE) # 10 samples in common
MB13 <- BuildMultiBlock(c1,c3,ignore.names = FALSE) # 5 samples in common
MB13b <- BuildMultiBlock(c1,c3,ignore.names = TRUE) # Blocks were appended
                                                    # regardless of their sample
                                                    # names. Sample names were
                                                    # replaced by integers.
# (Not run) The following code does not work because block sizes are different
# and ignore.names = TRUE.
#MB14 <- BuildMultiBlock(c1,c4,ignore.names = TRUE)
```

If we don't need to verify the sample correspondence across blocks, we can use
`ignore.names = TRUE`. However, in case the sample size is different across
blocks, the `MultiBlock` will not be built (as the sample correspondence does
not actually exist). It is possible to overrid this situation with 
`ignore.size = TRUE`, but the resulting `MultiBlock` is **not compatible** for 
`ComDim` analyses.

```{r MBsize, echo = TRUE, eval = FALSE}
# Option A (ignore.names = FALSE) : Only common samples across blocks are kept.
MB14c <- BuildMultiBlock(c1,c4,ignore.names = FALSE)
# Option B (ignore.names = TRUE, ignore.size = TRUE):
#    Blocks were appended regardless of their sample names and sizes.
#    All samples were kept.
MB14b <- BuildMultiBlock(c1,c4,ignore.names = TRUE, ignore.size = TRUE)
# MB14b is not compatible with ComDim
getSampleNames(MB14b)
# A particularity of this MB14b is that it has one 'Samples' vector per block.
```


## Splitting blocks by the Batch criterion. {#splitRW}

We can consider split one or more blocks according to the Batch criterion, in 
order to examine whether this factor contains is a relevant source of 
information in our dataset. This is performed with the `SplitRW()` function.
We call this split strategy **Replicate-Wise** (RW).

```{r Split, echo = TRUE, eval = FALSE}
# Build the MultiBlock
x1 = MultiBlock(Samples = 1:10,
                Data = list(x1 = matrix(rnorm(500),10,50)),
                Variables = list(x1 = 1:50),
                Batch = list(x1 = rep('Batch1',10)))
x2 = MultiBlock(Samples = c(1:10,1:10,1:10),
                Data = list(x2 = matrix(rnorm(2400),30,80)),
                Variables = list(x2 = 1:80),
                Batch = list(x2 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
                )
x1x2 <- BuildMultiBlock(x1,x2, ignore.names = TRUE, ignore.size = TRUE)
# Proceed with the split.
rw <- SplitRW(x1x2) # SplitRW looks specifically to the Batch information.
```

More information about the reasoning behind this analytical strategy can be 
consulted here: [**https://doi.org/10.1016/j.chemolab.2021.104422**]
(https://doi.org/10.1016/j.chemolab.2021.104422).

In case there exist sample correspondence although the sample names do not match
across blocks, we can force the MultiBlock split with the argument
`checkSampleCorrespondence = FALSE`.

```{r Split2, echo = TRUE, eval = FALSE}
# Build the MultiBlock
x4 = MultiBlock(Samples = c(1:30),
                Data = list(x4 = matrix(rnorm(2400),30,80)),
                Variables = list(x4 = 1:80),
                Batch = list(x4 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
)
x1x4 <- BuildMultiBlock(x1,x4, ignore.names = TRUE, ignore.size = TRUE)
  # Sample names in block x1 go from 1 to 10.
  # Sample names in block x4 go from 1 to 10 (batch1), 11 to 20 (batch2),
  # and 21 to 30 (batch3). Despite the sample names does not batch, samples
  # have correspondence across blocks. Thus, we need to impose **not to**
  # check the sample names. This is done with checkSampleCorrespondence = FALSE.
# Proceed with the split.
rw2 <- SplitRW(x1x4, checkSampleCorrespondence = FALSE)
```

With `checkSampleCorrespondence = TRUE`, only the common samples across blocks 
are kept.

```{r Split3, echo = TRUE, eval = FALSE}
# Build the MultiBlock
x5 = MultiBlock(Samples = c(1:10, 5:14, 1:10),
                Data = list(x5 = matrix(rnorm(2400),30,80)),
                Variables = list(x5 = 1:80),
                Batch = list(x5 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
)
x1x5 <- BuildMultiBlock(x1,x5, ignore.size = TRUE, ignore.names = TRUE)
# Proceed with the split.
rw3 <- SplitRW(x1x5, checkSampleCorrespondence = TRUE)
```
