---
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE, cache=FALSE}
library(R.ComDim)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 5,
  fig.height = 5,
  message = FALSE
)
```

```{r, echo = FALSE, eval = TRUE}
library(R.ComDim)
```

# Create a MultiBlock {#create}

## Option 1: for small MultiBlocks {#option1}

In order to use the ComDim algorithm, the data blocks need to be
combined into a `MultiBlock` object.

`MultiBlock` has 5 fields: `Samples`, `Data`, `Variables`, `Metadata`
and `Batch`. The easiest way to create a `MultiBlock` is by using the
function `MultiBlock()` as below:

```{r create, echo = TRUE, eval = TRUE}
set.seed(10)
b1 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
b2 = as.data.frame(matrix(rnorm(800),10,80)) # 10 rows and 80 columns
b2[c(2,3,4),c(5,7,8)] <- NA # Making some data missing, just for fun.
rownames(b2) <- LETTERS[5:14] # Adding some sample names

MB = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                s2 = 1:ncol(b2)))
str(MB)
```

With the code above, the `MultiBlock` MB containing 2 blocks was built.
As shown, the provided data blocks support the format `matrix` and
`data.frame`.

When building a `MultiBlock`, only the field `Data` is mandatory.
`Samples` and `Variables` are filled in automatically when omitted.

```{r create2, echo = TRUE, eval = TRUE}
b3 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
b4 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
MB2 = MultiBlock(Data = list(Block1 = b3, Block2 = b4))
str(MB2)
```

### `Batch` and `metadata`

You can pass additional sample metadata into a `MultiBlock` object,
either through the field `Batch` or through the field `metadata`. Within
each block, `Batch` data are vectors, while `metadata` are data.frames.

```{r create3, echo = TRUE, eval = TRUE}
MB3 = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                 s2 = 1:ncol(b2)),
                Batch = list(s1 = rep('Batch1',10),
                             s2 = c(rep('Batch2',5),rep('Batch3',5))))
str(MB3)
```

The `Batch` field can directly be added when the MultiBlock is created,
but the `metadata` needs to be created in a second step using the
`AddMetadata()` function. *Note: While ComDim analyses exclusively
utilize the field `Data`, retaining this metadata is often essential for
maintaining data consistency and traceability throughout your workflow.*

Another difference between `Batch` and `metadata` is that `Batch` refers
to the `MultiBlock` as a whole and `metadata` offers more granularity by
allowing for block-specific assignments. In the given example below, the
`metadata` contains two variables (`x1` and `x2`) and only apply to
`Block1`.

```{r AddMetadata, echo = TRUE, eval = TRUE}
b_mat1 <- matrix(rnorm(100), 10, 10)
b_mat2 <- matrix(rnorm(100), 10, 10)
MB4 <- MultiBlock(Data = list(Block1 = b_mat1, Block2 = b_mat2))

MB4_new <- AddMetadata(MB4,
                  block = 'Block1', # The block name
                  metadata = data.frame(x1 = c(rep(0,5),rep(1,5)),
                                        x2 = c(rep(1,3),rep(2,4),rep(3,3))))
str(MB4_new) # The MultiBlock before adding the metadata
```

`Metadata` can also be linked to several blocks at once as shown below.

```{r AddMetadata2, echo = TRUE, eval = TRUE}
MB4_new2 <- AddMetadata(MB4,
                  block = c('Block1','Block2'), # The block name
                  metadata = data.frame(x1 = c(rep(0,5),rep(1,5)),
                                        x2 = c(rep(1,3),rep(2,4),rep(3,3))))
str(MB4_new2) # The MultiBlock before adding the metadata
```

The last difference between `Batch` and `metadata` is that the `Batch`
field can be used to split the `MultiBlock` into smaller blocks, but the
`metadata` cannot. This will be later explained in Section
\@ref(splitRW).

## Option 2: for large MultiBlocks {#option2}

It is possible to create a `MultiBlock` by combining existing
`MultiBlock` objects. In this case, we pass them directly inside the
`Data` argument of `MultiBlock()`. Raw matrices and existing
`MultiBlock` objects can be mixed freely:

```{r create_option2, echo = TRUE, eval = TRUE}
MB5 <- MultiBlock(Data = list(b1 = b1, b2 = b2, MB = MB),
                 ignore.names = TRUE)
blockNames(MB5)
```

## Option 3: from SummarizedExperiment or MultiAssayExperiment objects {#option3}

`MultiBlock`s can also be created from `SummarizedExperiment` or
`MultiAssayExperiment` objects, which are pretty common format nowadays
in multi-omics studies. In the same way, `MultiBlocks` can be converted
back to `MultiAssayExperiments`.

```{r SummarizedExperiment, echo = TRUE, eval = FALSE}
library(MultiAssayExperiment)
data(miniACC)
library(SummarizedExperiment)
data(airway, package="airway")

MB1 <- SummarizedExperiment2MultiBlock(airway,
                                       colData_samplenames = 'Run',
                                       Batch = NULL)
MB2 <- MultiAssayExperiment2MultiBlock(miniACC,
                                       colData_samplenames = 'patientID',
                                       Batch = NULL)
se2 <-  MultiBlock2MultiAssayExperiment(MB2, MSEmetadata = NULL)
```

## Sample correspondence across blocks {#correspondence}

**Sample correspondence** refers to the alignment of samples across
different `Data` blocks. This ensures that the i-th row in one block
represents the same biological sample as the i-th row in all other
blocks. Most multi-omics frameworks require this correspondence. In
**R.ComDim**, the `MultiBlock()` function automates the verification of
sample correspondence. By default `(ignore.names = FALSE)`, the function
checks for consistent sample names across blocks and will automatically
reorder or discard samples to ensure a perfect match.

```{r correspondence1, echo = TRUE, eval = TRUE}
c1 = matrix(1:500,10,50) # 10 rows and 50 columns
c2 = matrix(500:1,10,50) # 10 rows and 50 columns
c3 = matrix(501:1000,10,50) # 10 rows and 50 columns
c4 = matrix(1:1000,20,50) # 20 rows and 50 columns
rownames(c1) <- paste0('c',6:15)
rownames(c2) <- paste0('c',1:10)
rownames(c3) <- paste0('c',10:1)
rownames(c4) <- paste0('c',1:20)
# With ignore.names = FALSE, only common samples across blocks are kept.
# Samples will be resorted if needed.
MB12 <- MultiBlock(Data = list(c1 = c1, c2 = c2), 
                   ignore.names = FALSE)
# 5 samples in common: c6, c7, c8, c9, c10

MB13 <- MultiBlock(Data = list(c1 = c1, c3 = c3),
                   ignore.names = FALSE)
# 5 samples in common: c6, c7, c8, c9, c10

MB23 <- MultiBlock(Data = list(c2 = c2, c3 = c3),
                   ignore.names = FALSE)
# 10 samples in common, samples were resorted.
# New order was c1, c2, c3, c4, c5, c6, c7, c8, c9, c10
```

```{r correspondence2, echo = TRUE, eval = TRUE}
MB13b <- MultiBlock(Data = list(c1 = c1, c3 = c3), ignore.names = TRUE)
MB13b@Samples
```

In the case above, the number of samples is 10 and not 5 because the
sample correspondence was not checked `(ignore.names = TRUE)`. Blocks
were appended regardless of their sample names. Sample names were
replaced by integers.

If we don't need to verify the sample correspondence across blocks, we
can use `ignore.names = TRUE`. However, in case the sample size is
different across blocks, the `MultiBlock` will not be built (as the
sample correspondence does not actually exist!).

```{r correspondence3, echo = TRUE, eval = FALSE}
MB14 <- MultiBlock(Data = list(c1 = c1, c4 = c4), ignore.names = TRUE)
# Error in MultiBlock(Data = list(c1 = c1, c4 = c4), ignore.names = TRUE) : 
# Blocks do not have the same number of rows. Provide sample names for all blocks
# to enable matching, or set 'ignore.size = TRUE'.
```

It is possible to override this situation with `ignore.size = TRUE`, but
the resulting `MultiBlock` is **not compatible** for `ComDim` analyses.

```{r MBsize, echo = TRUE, eval = TRUE}
# Option A (ignore.names = FALSE) : Only common samples across blocks are kept.
MB14A <- MultiBlock(Data = list(c1 = c1, c4 = c4), ignore.names = FALSE)
rownames(c1)
rownames(c4)
MB14A@Samples

# Option B (ignore.names = TRUE, ignore.size = TRUE):
# Blocks were appended regardless of their sample names and sizes.
# All samples were kept.
MB14B <- MultiBlock(Data = list(c1 = c1, c4 = c4), ignore.names = TRUE, ignore.size = TRUE)
# MB14B is not compatible with ComDim
MB14B@Samples
# A particularity of this MB14B is that it has one 'Samples' vector per block.
```

## Splitting blocks by `Batch` {#splitRW}

We can consider split one or more blocks by `Batch`, in order to examine
whether this factor contains is a relevant source of information in our
dataset. This is performed with the `SplitRW()` function. We call this
split strategy **Replicate-Wise** (RW).

```{r Split1, echo = TRUE, eval = TRUE}
# Build the MultiBlock
x1 = MultiBlock(Samples = 1:10,
                Data = list(x1 = matrix(rnorm(500),10,50)),
                Variables = list(x1 = 1:50),
                Batch = list(x1 = rep('Batch1',10)))
x2 = MultiBlock(Samples = c(1:10,1:10,1:10),
                Data = list(x2 = matrix(rnorm(2400),30,80)),
                Variables = list(x2 = 1:80),
                Batch = list(x2 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
                )
x1x2 <- MultiBlock(Data = c(x1@Data, x2@Data),
                   Batch = c(x1@Batch, x2@Batch),
                   ignore.names = TRUE, ignore.size = TRUE)
```

In the `MultiBlock` above, each studyset contains 10 samples. However,
the first block has 1 replicate (1 "Batch") per sample and the second
block has 3 replicates (3 "Batches") per sample. We can therefore
recover the sample correspondence by splitting the second block into
three smaller blocks of 10 samples each. The resulting multiblock will
consist of 4 (1+3) blocks.

```{r Split2, echo = TRUE, eval = TRUE}
# Proceed with the split.
rw <- SplitRW(x1x2) # SplitRW looks specifically to the Batch information.
str(rw)
```

When `splitRW()` is applied, a data.frame with the sample correspondence
across the resulting batches is given in a message, as shown above.

More information about the reasoning behind this analytical strategy can
be consulted here:
[[**https://doi.org/10.1016/j.chemolab.2021.104422**](https://doi.org/10.1016/j.chemolab.2021.104422){.uri}]
(<https://doi.org/10.1016/j.chemolab.2021.104422>).

In case we know that there exist sample correspondence although the
sample names do not match across blocks, we can force the `MultiBlock`
split anyways with the argument `checkSampleCorrespondence = FALSE`.

```{r Split3, echo = TRUE, eval = TRUE}
# Build the MultiBlock
x4 = MultiBlock(Samples = c(1:30),
                Data = list(x4 = matrix(rnorm(2400),30,80)),
                Variables = list(x4 = 1:80),
                Batch = list(x4 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
)
x1x4 <- MultiBlock(Data = c(x1@Data, x4@Data),
                   Batch = c(x1@Batch, x4@Batch),
                   ignore.names = TRUE, ignore.size = TRUE)
  # Sample names in block x1 go from 1 to 10.
  # Sample names in block x4 go from 1 to 10 (batch1), 11 to 20 (batch2),
  # and 21 to 30 (batch3). Even though the sample names do not match,
  # we know that samples have correspondence across blocks.
  # We therefore need to skip sample name checking with
  # checkSampleCorrespondence = FALSE.

# Proceed with the split.
rw2 <- SplitRW(x1x4, checkSampleCorrespondence = FALSE)
```

With `checkSampleCorrespondence = TRUE`, only the common samples across
blocks are kept.

```{r Split4, echo = TRUE, eval = TRUE}
# Build the MultiBlock
x5 = MultiBlock(Samples = c(1:10, 5:14, 1:10),
                Data = list(x5 = matrix(rnorm(2400),30,80)),
                Variables = list(x5 = 1:80),
                Batch = list(x5 = c(rep('Batch1',10),
                                    rep('Batch2',10),
                                    rep('Batch3',10)))
)

x1x5 <- MultiBlock(Data = list(x1 = x1, x5 = x5),
                   ignore.names = TRUE, ignore.size = TRUE)
# Proceed with the split.
# Samples 5-10 are common to all batches (x1: 1-10; x5 Batch1: 1-10;
# x5 Batch2: 5-14; x5 Batch3: 1-10), so rw3 will contain 6 samples.
rw3 <- SplitRW(x1x5, checkSampleCorrespondence = TRUE)
```
