```{r include=FALSE, cache=FALSE}
library(R.ComDim)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 5,
  fig.height = 5,
  message = FALSE
)
```

```{r, echo = FALSE, eval = TRUE}
library(R.ComDim)

b1 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
b2 = as.data.frame(matrix(rnorm(800),10,80)) # 10 rows and 80 columns
b2[c(2,3,4),c(5,7,8)] <- NA # Making some data missing, just for fun.
rownames(b2) <- LETTERS[5:14] # Adding some sample names
b3 = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                s2 = 1:ncol(b2)))
MB <- MultiBlock(Data = c(list(b1 = b1, b2 = b2), b3@Data),
                 ignore.names = TRUE)
```

# MultiBlock data handling {#handling}

## Inspecting the data

The slots of a `MultiBlock` object can be accessed directly with `@`. The five
slots are `@Samples`, `@Data`, `@Variables`, `@Batch`, and `@Metadata`.

```{r handling, echo = TRUE, eval = TRUE}
MB@Samples            # Sample names (vector, or list when blocks differ)
MB@Variables          # Variable names for all blocks (named list)
MB@Variables[[2]]     # Variable names for the second block
MB@Variables[["s2"]]  # Variable names for a block by name
names(MB@Data)        # Block names
names(MB@Batch)       # Blocks that have batch labels
names(MB@Metadata)    # Blocks that have metadata
ncol(MB)
nrow(MB)
MB2 <- FilterSamplesMultiBlock(MB, c(1:9)) # To create a MultiBlock with
 # a sample subset. In this case, we selected samples 1-9.
```

Or we can simply see all the content with `str()` function.

If needed, everything in the `MultiBlock` can be renamed by direct slot
assignment.

```{r rename, echo = TRUE, eval = TRUE}
# Rename blocks (update @Data and @Variables names together)
names(MB@Data)      <- paste("X", 1:4, sep = '')
names(MB@Variables) <- paste("X", 1:4, sep = '')
names(MB@Data)

# Rename samples
MB@Samples <- LETTERS[1:10]
MB@Samples

# Rename variables in a specific block
MB@Variables[["X1"]] <- paste("vars", 1:50, sep = '')
MB@Variables[["X1"]]
```

## Data pre-processing {#processing}

The `R.ComDim` package contains functions to easily apply some data
transformations. This can result in handy since we there is no need to build
 the `MultiBlock` every time a new data transformation is tested.
```{r processing1, echo = TRUE, eval = TRUE}
## NA removal
# Block names are X1–X4 after the renaming step above.
MB@Data$X2[c(2,3,5),c(1,2,3)] <- NA # Add some NAs to block X2
allMB <- NAInfRemoveMultiBlock(MB, na.method = 'none', minfrac = 0.2)
# Variables containing more than 20% of NAs will be discarded

## Data normalization
allMB <- NormalizeMultiBlock(allMB, method = 'norm')
# MB is normalized (mean-center and divided by the block norm)
```

The `R.ComDim` package has been designed with flexibility in mind, and allows
the user to apply custom data transformations. For instance, in the code below,
all variables with values lower than 5% of the most intense value are discarded.

```{r processing2, echo = TRUE, eval = TRUE}
# MB is converted to matrix to calculate the max value.
maxMB <- max(MultiBlock2Matrix(allMB), na.rm = TRUE)
# Variables are filtered with the ProcessMultiBlock() function.
allMB <- ProcessMultiBlock(allMB,
  FUN.SelectVars = function(x) {apply(x,2,max) > maxMB * 0.05})
```

`ProcessMultiBlock()` can also apply a custom transformation to the data (`FUN`),
or remove entire blocks based on a criterion (`FUN.SelectBlocks`). The three
arguments can be combined; when they are, the order of execution is
`FUN` → `FUN.SelectVars` → `FUN.SelectBlocks`. The `blocks` argument restricts
processing to a subset of blocks, while `vars` allows explicit variable
selection per block.

```{r processing3, echo = TRUE, eval = TRUE}
# Apply a custom transformation: express each block as a percentage of its maximum.
BY100 <- function(x) { 100 * (x / max(x, na.rm = TRUE)) }
allMB <- ProcessMultiBlock(allMB, FUN = BY100)

# Process only specific blocks (by name or by index):
allMB <- ProcessMultiBlock(allMB, blocks = c('X1', 'X2'), FUN = BY100)

# Keep only a specific set of variables in each processed block.
# 'vars' must be a list with one element per block listed in 'blocks'.
allMB <- ProcessMultiBlock(allMB,
  blocks = c('X1', 'X2'),
  vars   = list(1:20, 1:40))   # first 20 vars of X1, first 40 vars of X2

# Remove blocks whose Frobenius norm is below a threshold
# (FUN.SelectBlocks should return TRUE to indicate the block passes the criterion;
#  blocks for which it returns FALSE are removed):
norm_threshold <- function(x) { norm(x, type = 'F') > 50 }
allMB <- ProcessMultiBlock(allMB, FUN.SelectBlocks = norm_threshold)
```

### Normalization methods {#normalization}

`NormalizeMultiBlock()` provides several normalization strategies. The default
(`'norm'`) mean-centers each block and divides by its Frobenius norm — this is
the standard preparation before running ComDim. Other methods are available for
exploratory work or when a specific pre-processing is required.

```{r processing4, echo = TRUE, eval = TRUE}
# Reset to NA-removed, un-normalized data so each method is applied independently.
allMB <- NAInfRemoveMultiBlock(MB, na.method = 'zero', minfrac = 0.2)

# 'norm' (default): mean-center + divide by Frobenius norm. Recommended for ComDim.
allMB_norm    <- NormalizeMultiBlock(allMB, method = 'norm')

# 'auto': autoscaling (mean-center + divide by SD per variable, i.e. unit variance).
allMB_auto    <- NormalizeMultiBlock(allMB, method = 'auto')

# 'mean': mean-centering only (no scaling).
allMB_mean    <- NormalizeMultiBlock(allMB, method = 'mean')

# 'pareto': mean-center + divide by sqrt(SD) per variable.
allMB_pareto  <- NormalizeMultiBlock(allMB, method = 'pareto')

# 'geometric': all values must be positive after the constant shift.
#   Random (or mean-centered) data can be negative, so shift each block to
#   make it strictly positive before calling NormalizeMultiBlock.
allMB_pos  <- ProcessMultiBlock(allMB, FUN = function(x) x - min(x, na.rm = TRUE) + 1)
allMB_geo  <- NormalizeMultiBlock(allMB_pos, method = 'geometric')

# 'ranknorm': rank-based inverse normal transform (Blom transform by default).
allMB_rn      <- NormalizeMultiBlock(allMB, method = 'ranknorm')

# 'none': no transformation (still converts Inf to NA when infinite.as.NA = TRUE).
allMB_none    <- NormalizeMultiBlock(allMB, method = 'none', infinite.as.NA = TRUE)

# Normalize only selected blocks:
allMB         <- NormalizeMultiBlock(allMB, blocks = c('X1', 'X3'), method = 'auto')
```

### Adding batch and metadata information {#metadata}

Batch labels and per-sample metadata can be attached to any block with
`AddMetadata()`. This information is used by `SplitRW()` (Chapter \@ref(splitRW))
and is preserved throughout the ComDim workflow.

```{r processing5, echo = TRUE, eval = TRUE}
b1 <- matrix(rnorm(500), 10, 50)   # 10 samples, 50 variables
b2 <- matrix(rnorm(800), 10, 80)   # 10 samples, 80 variables
mb <- MultiBlock(Data = list(b1 = b1, b2 = b2))

# Add batch labels (must be a numeric vector, one value per sample):
batch_labels <- c(1, 1, 1, 1, 2, 2, 2, 2, 3, 3)
mb <- AddMetadata(mb, block = 'b1', batches = batch_labels)
mb <- AddMetadata(mb, block = 'b2', batches = batch_labels)

# Add sample metadata (must be a data.frame with one row per sample):
meta <- data.frame(
  condition = rep(c('Control', 'Treatment'), 5),
  time      = c(0, 0, 6, 6, 12, 12, 24, 24, 48, 48)
)
mb <- AddMetadata(mb, block = 'b1', metadata = meta)

# Batch labels and metadata can be set together in one call:
mb <- AddMetadata(mb, block = 'b2', batches = batch_labels, metadata = meta)

# Multiple blocks can be updated at once if they share the same labels:
mb <- AddMetadata(mb, block = c('b1', 'b2'), batches = batch_labels)
```
