```{r include=FALSE, cache=FALSE}
library(R.ComDim)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 5,
  fig.height = 5,
  message = FALSE
)
```

```{r, echo = FALSE, eval = TRUE}
library(R.ComDim)

b1 = matrix(rnorm(500),10,50) # 10 rows and 50 columns
b2 = as.data.frame(matrix(rnorm(800),10,80)) # 10 rows and 80 columns
b2[c(2,3,4),c(5,7,8)] <- NA # Making some data missing, just for fun.
rownames(b2) <- LETTERS[5:14] # Adding some sample names
b3 = MultiBlock(Samples = 1:10,
                Data = list(s1 = b1, s2 = b2),
                Variables = list(s1 = 1:ncol(b1),
                                s2 = 1:ncol(b2)))
MB <- BuildMultiBlock(b1,b2,b3,ignore.names = TRUE)
```

# MultiBlock data handling {#handling}

## Inspecting the data

We can easily read the content of a MultiBlock object with the following
functions:

```{r handling, echo = TRUE, eval = TRUE}
getSampleNames(MB)
getVariableNames(MB)
getVariableNames(MB, block = 2)
getBlockNames(MB)
getBlockNames(MB, "Batch")
getBlockNames(MB, "Metadata")
ncolMultiBlock(MB)
nrowMultiBlock(MB)
MB2<- FilterSamplesMultiBlock(MB, c(1:9)) # To create a MultiBlock with 
 # a sample subset. In this case, we selected samples 1-9.
```

Or we can simply see all the content with `str()` function.

If needed, everything in the MB can be renamed.
```{r rename, echo = TRUE, eval = FALSE}
MB <- setBlockNames(MB, paste("X", 1:4, sep = '')) # The blocks
getBlockNames(MB)
MB <- setSampleNames(MB, LETTERS[1:10]) # The samples
getSampleNames(MB)
MB <- setVariableNames(MB, paste("vars", 1:50, sep = ''), 1)
getVariableNames(MB, 1)
```

## Data pre-processing {#processing}

The `R.ComDim` package contains functions to easily apply some data
transformations. This can result in handy since we there is no need to build
 the `MultiBlock` every time a new data transformation is tested.
```{r processing1, echo = TRUE, eval = FALSE}
## NA removal
# We first add some NAs to the MultiBlock.
MB@Data$X2[c(2,3,5),c(1,2,3)] <- NA # Add some NAs
allMB <- NARemoveMultiBlock(MB, method = 'none', minfrac = 0.2)
# Variables containing more than 20% of NAs will be discarded

## Data normalization
allMB <- NormalizeMultiBlock(allMB, method = 'norm')
# MB is normalized (mean-center and divided by the block norm)
```

The `R.ComDim` package has been writted with the idea to result very flexible 
and, as such, it allows the user to apply custom data transformations. For 
instance, in the code below, all variables with values lower than the 5% of the 
most intense value are discarded.

```{r processing2, echo = TRUE, eval = FALSE}
# MB is converted to matrix to calculate the max value.
maxMB <- max(MultiBlock2matrix(allMB), na.rm = TRUE) 
# Variables are filtered with the ProcessMultiBlock() function.
allMB <- ProcessMultiBlock(allMB,
  FUN.SelectVars = function(x) {apply(x,2,max) > maxMB * 0.05})
```
